The Impact of Accretion Disk Winds on the Optical Spectra of Cataclysmic Variables

### 2014-12-03 draft.tex
updated some figures

\begin{abstract}
Many high-state non-magnetic cataclysmic variables (CVs) exhibit
blue-shifted absorption or P-Cygni profiles in ultraviolet
(UV) resonance lines, such as C~{\sc iv}~1550\AA. These features
imply the existence of powerful accretion disk winds in
CVs, and considerable effort has gone into modelling UV line formation
in such outflows. By contrast, much less attention has been paid to
the potential impact of these disk winds on the {\em optical} spectra
of CVs. Here, we use an improved version of our Monte Carlo ionization
and radiative transfer code to investigate whether disk wind models that
produce realistic UV line profiles are also likely to generate
observationally significant recombination line and continuum emission
in the optical waveband. In particular, we test whether outflows may 
be responsible for the single-peaked emission line profiles often seen
in high-state CVs and for the weakness of the Balmer absorption edge
(relative to simple models of optically thick accretion disks).
We find that a standard disk wind model that is successful in
reproducing the UV spectra of CVs does leave a noticeable imprint on
the optical spectrum as well, particularly for systems viewed at high
inclination. The strongest optical wind-formed recombinations lines are \ha\ and
\heiiopt\, although the outflow also produces emission in the Ly-$\alpha$
and \heiiuv\ UV lines. However, all of the optical lines produced by
this default wind model are double-peaked, and the effect of the wind
on the Balmer jump is modest. We carry out a limited exploration of
the disk wind parameter space, focusing on those parameters that affect the
density near the base of the wind. This shows that higher-density
outflows can produce both single-peaked emission lines and a
recombination continuum sufficient to fill in the Balmer
jump, although such models also tend to predict stronger-than-observed 
collisionally excited UV emission lines. Our main conclusion is that
disk wind emission is likely to have a significant impact on the
optical spectra of high-state CVs. 
{\bf NOTE JM: This abstract is a few lines longer than I would like- 
I think something punchier is required.}
\end{abstract}

@@

Much less is known about the effect of these outflows on the optical
spectra of high-state CVs. These spectra are typically characterized
by H and He emissions superposed on a blue continuum. In many
cases, and particularly in the SW~Sex subclass of NLs
\citep{HSK86,DR95}, these lines are single-peaked. This is contrary to
theoretical expectations for lines formed in accretion disks, which
are predicted to be double-peaked \citep{smak1981, hornemarsh1986}. 
{\em Low-state} CVs (dwarf novae in quiescence) can, in fact,
exhibit such double-peaked lines \citep{marshhorne1990}. 
{\bf NOTEJM: I've included a reference to a paper on IP Peg here
which shows the transition from double peaked lines in 
quiescence and a contrast to outburst spectra. I'm not sure
of general trends in DNe. A lot seem to have weak emission cores
embedded in absorption.}

@@

\subsection{Ionization and Excitation: ``Simple Atoms''}

@@

\begin{figure}
\centering
\includegraphics[width=0.5\textwidth]{figures/fig_caseb_tardis.eps}
\caption{
{\sl Top Panel:} Balmer decrements calculated with \py compared with analytic calculations
by Seaton (1959), showing good agreement when calculating level emissivities in Hydrogen
in the Case B limit.
(see Osterbrock 1989 for a discussion of this commonly used approximation).
{\sl Bottom Panel:}  a comparison of Helium I level populations (the most complex ion we currently 
treat as a macro-atom) between \py~and \tar~models. Considering the 
two codes use different atomic data and \textsc{Tardis,} unlike \textsc{Python,} currently has 
a complete treatment of collisions between radiatively forbidden transitions, the factor of 
$<2$ agreement is encouraging. 
{\bf NOTE CK: WHAT ARE THE PARTICULAR SETS OF PARAMETERS BEING ADOPTED HERE?}
}
\label{tests}
\end{figure}
\nocite{osterbrock}
\nocite{seaton1959}

@@

\subsection{Wind Geometry and Kinematics}

@@

The density at position $(r,z)$ in the wind, $\rho(r,z)$, is
calculated from the mass continuity equation, yielding
\begin{equation}
\rho(r,z) = \frac{r}{r_0} \frac{dr}{dr_0} \frac{\phi(r_0)}{v_z(r,z)}.
\label{density}
\end{equation}
Here, $\phi(r_O)$ is the mass-loss rate per unit area at $r_0$, 
and $v_z$ is the vertical velocity component. We take $\phi(r_0)$ to be
uniform across the wind-launching area of the disk and normalize it by
matching its integral to the user-specified total mass-loss rate,
$\dot{M}_{wind}$. 
{\bf NOTE CK: I'm still not sure I understand the formula
  SV93 use to define $\phi$, and hence that this description is
  actually strictly correct.... Not sure what to do about that
  though...}

\subsection{Sources and Sinks of Radiation}

@@

It is possible to include radiation from a boundary layer (BL) between
the disk and the WD. In \py, the BL is described as
a blackbody with a user-specified effective temperature and
luminosity. In the models presented here, we have followed LK02 in setting
the BL luminosity to zero. However, we have confirmed that the addition of a
BL with $L_{BL} = 0.5 L_{acc}$ and temperatures in the range $80{\rm
kK} \leq T_{BL} \leq 200{\rm kK}$ would not change any of our main
conclusions. 
{\bf NOTE CK: Is this true? Do we want to say anything else, e.g. that we'll
talk about this in a future paper? JM: This is true. Anything up to 200kK
still has optical lines. The He II behaviour is not monotnic. CIV drops
gradually and gets a little overionized- but this is a separate topic really.
The issue of clumping and BL and their affect on ionization could be a 
future paper.}

@@

\begin{figure}
\centering
\includegraphics[width=0.45\textwidth]{figures/sed_figure.eps}
\caption{
The spectral energy distribution (SED) of the blackbody accretion
disk used in the ionization calculation (black line), 
with two boundary layer blackbody spectra also plotted.
The positions of the Hydrogen and Helium ionization edges 
are marked with dashed vertical lines.
{\bf JM: do we want to show the stellar atmosphere spectrum here?
we know it looks a bit crap around the He II edge, so perhaps not...
 }
\label{sed}
\end{figure}

@@

\begin{table}
\centering
\begin{tabular}{p{3cm}p{4cm}}
Benchmark Model (Optimized for the UV Band) \\
\hline Free Parameters 	&	 Value \\ 
\hline \hline 
$M_{WD}$ 	 &	 $0.8 M_{\odot}$ \\ 
$R_{WD}$ 	 &	 $7\times10^{8}$cm\\ 
$T_{WD}$ 	 &	 $40,000$K\\
$M_{2}$ 	 &	 $???? M_{\odot}$ \\ 
$\dot{M}_{acc}$ 	 &	 $10^{-8}~M_{\odot}yr^{-1}$\\ 
$\dot{M}_{wind}$  &	$10^{-9}~M_{\odot}yr^{-1}$å\\ 
$r_{min}$ 	&	 $4 R_{WD}$\\ 
$r_{max}$ 	&	 $12 R_{WD}$ \\ 
$\theta_{min}$ 	&	 $20.0^{\circ}$ \\ 
$\theta_{max}$ 	&	 $65.0^{\circ}$ \\ 
$\gamma$ 	&	 $1$ \\ 
$v_{\infty}$ 	&	 $3v_{esc}$ \\ 
%%$R_v$ 	        &	 $7\times10^{10}$cm \\ 
$R_v$ 	        &	 $100 R_{WD}$ \\ 
$\alpha$ 	&	 $1.5$ \\
\end{tabular}
\centering
\caption{
Parameters used for the geometry and kinematics of the benchmark CV model.}
\label{wind_param}
\end{table}

\section{A Benchmark Disk Wind Model}

@@

\subsection{Physical Structure and Ionization State}

Figure~\ref{wind} shows the physical and ionization structure 
of the benchmark disk wind model. There is an obvious drop-off in density
and temperature with distance away from the disk, so any line
formation process that scales as $n_e^2$ -- i.e. recombination and
collisionally excited emission -- should be expected to operate
primarily in the dense base of the outflow. Moreover, a comparison of
the rotational and poloidal velocity fields shows that rotation
dominates in the near-disk regime, while outflow dominates further out
in the wind.

The ionization equation used in the ``simple atom'' approach used in
LK02 (see Section~\ref{BLA}) should be a reasonable approximation to
the photoionization equilibrium in the benchmark wind model. Even
though the macro-atom treatment of H and He does affect the 
computation of the overall ionization equilibrium, we would expect the
resulting ionization state of the wind to be quite similar to that
found in LK02. The bottom panels in Figure~\ref{BL} confirm that this
is the case. In particular, Helium is fully ionized
throughout most of the outflow, except for a small region near the
base of the wind, which is shielded from the photons produced by the
hot inner disk. In line with the results shown in LK02, we also find
that C\textsc{iv} is the dominant Carbon ion throughout the wind,
resulting in a substantial absorbing column across a large range of
velocities. As we shall see, this produces the broad, deep and
blue-shifted C\textsc{iv}~$\lambda 1550{\rm \AA}$ absorption line that
is usually the most prominent wind-formed feature in the UV spectra of
low-inclination nova-like CVs.

\subsection{Synthetic Spectra}

We begin by verifying that the benchmark model still produces UV
spectra that resemble those observed in CVs. We do expect this to be
the case, since the ionization state of the wind has not changed
significantly from that computed in LK02 (see Section~\ref{BLA}). The
left column of panels in Figure~\ref{spec} shows that this expectation
is met: all of the strong metal resonance
lines - notably N\textsc{v}~$\lambda 1240{\rm \AA}$,
Si\textsc{iv}~$\lambda 1400{\rm \AA}$ and C\textsc{iv}~$\lambda
1550{\rm \AA}$ - are present and exhibit clear P-Cygni profiles
at intermediate inclinations. In addition, however, we now also find
that the wind produces significant Lyman-$\alpha$ and
He~\textsc{ii}~$\lambda1640{\rm \AA}$ emission lines.

@@

Figure~\ref{spec} shows the corresponding optical spectra produced for
the benchmark model, and these do exhibit some emission lines
associated with H and He. Since the emissivity of these recombination 
features scales as $n_e^2$, they are formed almost entirely in the 
dense base of the wind, just above the accretion disk. Here, the
velocity field of the wind is still dominated by rotation, rather than
outflow, which accounts for the double-peaked shape of the lines. In
principle, lines formed in this region can still be single peaked,
since the existence of a poloidal velocity {\em gradient} changes the
local escape probabilities (MC96). However, as
demonstrated explicitly in Section~\ref{BLA}, the line opacity in our
benchmark model is not high enough for this radiative transfer effect
to dominate the line shapes.

@@

\section{The Dependence of Wind Features on Outflow Parameters}

The results of Section~\ref{BLA} show that wind models that reproduce
the UV spectral features of CVs are likely to leave a clear imprint on
the optical spectrum. It is therefore important to understand the
sensitivity of optical (and UV) wind-formed features to the basic
properties of the outflow. In particular, there are two key questions
we would like to address: (i) under what conditions can disk winds
produce {\em single-peaked} Balmer emission lines (MC96); (ii) can the wind-formed recombination continuum completely
fill in the disk's Balmer absorption edge for reasonable outflow
parameters?

@@

The simplest way to simultaneously affect the density in the wind as
well as the velocity gradients is by modifying the poloidal velocity
law. We have therefore conducted a exploratory sensitivity study in
which we focus on just two kinematic variables (Section~\ref{BLA}):
\begin{itemize}
 	\item the acceleration length, $R_v$, which controls the
        distance over which the wind accelerates to $\frac{1}{2}v_{\infty}$;
 	\item the acceleration exponent, $\alpha$, which controls the rate 
 	at which the poloidal velocity changes near $R_v$.
\end{itemize} 
The general behaviour we might expect is that outflows with denser
regions near the wind base -- i.e. winds with larger $R_{v}$ and/or
larger $\alpha$ -- will produce stronger optical emission signatures. 
However, this behaviour may be moderated by the effect of the increasing
optical depth through this region.

\subsection{Optical Wind Signatures}

Figures~\ref{halpha} and~\ref{heii} show how the H$\alpha$ and 
He~{\sc ii}~$\lambda 
4686$ lines change with the kinematics of the wind for an inclination
of $80^\circ$. The width and strength of He~{\sc ii} clearly depend on
both $R_v$ and $\alpha$, but the overall line profile shape is always
double-peaked. This contrasts with the behaviour of H$\alpha$, which
becomes single-peaked in our densest models. The transition from
double-peaked to single-peaked occurs when the line becomes optically
thick. This reproduces the behaviour predicted by \cite{MC96}, who
showed that the poloidal velocity gradient in the wind can modify the
shape of optically thick emission lines even in the near-disk regime,
where the poloidal velocity itself is much smaller than the rotational
velocity. This has the effect of suppressing the line wings and
can therefore lead to the production of a single-peaked line profile
shape.

@@

Two other features are worth noting in this part of the
spectrum. First, the collisionally excited Ca~{\sc ii} H\&K 
emission lines at 3969\AA 
and (especially) 3934\AA become quite prominent in our densest
models. Second, {\em all} of our models predict a detectable
He~\textsc{ii} recombination line at 3202\AA. This is the Helium
equivalent of Paschen~$\beta$ and should be expected in all systems that
feature a strong He~\textsc{ii}~4686\AA\ line (the Helium
equivalent of Paschen~$\alpha$). This line is unfamiliar observationally,
but only because it lies bluewards of the atmospheric cut-off, but
also redwards of most ultraviolet spectra. 
{\bf JM: should we reference something for this paper here}

\begin{figure} %fullpage
\mbox{
\subfigure{\includegraphics[width=0.5\textwidth]{figures/3by3_grid_alpha.eps}}
\quad
\subfigure{\includegraphics[width=0.5\textwidth]{figures/3by3_grid_4686.eps}}   
}
\caption{
Line profiles plotted in veliocity space 
for \ha\ (left) and \heiiopt\ (right) with varying kinematic 
properties, computed for an inclination of $80^\circ$.
The benchmark model is shown in the top left panel, and the improved optical
model described in section 6 is highlighted in bold. 
The parameter that has been altered from the benchmark model 
is marked in each column and row. Models towards the bottom right
are have more slowly accelerating winds and denser wind bases. }
\label{halpha}
\end{figure} %fullpage

\begin{figure} %fullpage
\centering
\includegraphics[width=0.8\textwidth]{figures/3by3_grid_balmer.eps}  
%%\includegraphics[width=0.5\textwidth]{figures/test.eps}
\caption{
Line profiles for \civ\ (left) and \heiiuv\  (right) with varying kinematic 
properties, computed for an inclination of $80^\circ$. 
The panels are organised in the same way as figure~\ref{halpha}.
% The panels are ordered in the same way as figure~\ref{halpha}. In general, increasing densities
% lead to a Balmer jump in emission, although as in the case of H$\alpha$ there is a transition
% when the wind becomes a neutral absorbing column.
}
\label{uvlines}
\end{figure} %fullpage

\subsection{Ultraviolet Wind Signatures}

The dependence of the ultraviolet C~{\sc iv} and He~{\sc ii} lines on
the same kinematic wind parameters is illustrated in Figures~\ref{BLA}
and~\ref{BLA}. As before, we focus on the predicted behaviour at $i =
80^\circ$. In our benchmark model, the C~{\sc iv} emission
line seen at high inclinations is almost entirely produced via
resonance scattering. The angle-averaged flux integrated over the
entire line profile - accounting correctly for the blue-shifted
absorption component seen at low inclinations - would be
approximately zero. However, as we increase the density and extent of
the slow-moving part of the outflow near the disk plane, collisionally
excited emission starts to contribute significantly to the C~{\sc iv}
line profiles. This explains the much stronger line-to-continuum
contrast seen in the $\alpha = 1.5$, $R_v = 2\times 10^{11}$~cm model,
for example. For any given inclination, the trend is not necessarily
monotonic, however, because the escape probabilities from the
line-forming region can be highly anisotropic.  
{\bf CK: I *think* this is the basic reason, but (a) can you check
  this, James; (b) is it enough to just point this out?; (c) do we
  understand why CIV is single-peaked but HeII, for example, is double-peaked?}

@@

\begin{figure} %fullpage
\mbox{
\subfigure{\includegraphics[width=0.5\textwidth]{figures/3by3_grid_uv.eps}}
\quad
\subfigure{\includegraphics[width=0.5\textwidth]{figures/3by3_grid_1640.eps}}   
}
%%\includegraphics[width=0.5\textwidth]{figures/test.eps}
\caption{
Line profiles for \civ\ (left) and \heiiuv\  (right) with varying kinematic 
properties, computed for an inclination of $80^\circ$. 
The panels are organised in the same way as figure~\ref{halpha}.
% The panels are ordered in the same way as figure~\ref{halpha}. In general, increasing densities
% lead to a Balmer jump in emission, although as in the case of H$\alpha$ there is a transition
% when the wind becomes a neutral absorbing column.
}
\label{uvlines}
\end{figure}

@@

However, perhaps the most serious problem for this optically optimized
model is associated with its predicted UV spectra. In order to
generate strong optical wind signatures, we have adopted wind
parameters that lead to very high densities at the base of the wind
($n_e\sim10^{13}-10^{14}$ in the main emission region). This produces
the desired optical recombination emission, but also increases the
role of collisional excitation in the formation of the UV resonance
lines. This explains the pronounced increase in the emission component 
of the C\textsc{iv} 1550\AA resonance line, for example, relative to
what was seen in the benchmark model (compare Figures~\ref{BLA} and
\ref{BLA}). The strength of this component in the revised model --
especially at low and intermediate inclinations -- is probably
somewhat too high to be consistent with UV observations of high-state
CVs (REF).

>>>>

\begin{abstract}
Many high-state non-magnetic cataclysmic variables (CVs) exhibit
blue-shifted absorption or P-Cygni profiles in ultraviolet
(UV) resonance lines, such as C~{\sc iv}~1550\AA. These features
imply the existence of powerful accretion disk winds in
CVs, and considerable effort has gone into modelling UV line formation
in such outflows. By contrast, much less attention has been paid to
the potential impact of these disk winds on the {\em optical} spectra
of CVs. Here, we use an improved version of our Monte Carlo ionization
and radiative transfer code to investigate whether disk wind models that
produce realistic UV line profiles are also likely to generate
observationally significant recombination line and continuum emission
in the optical waveband. In particular, we test whether outflows may 
be responsible for the single-peaked emission line profiles often seen
in high-state CVs and for the weakness of the Balmer absorption edge
(relative to simple models of optically thick accretion disks).
We find that a standard disk wind model that is successful in
reproducing the UV spectra of CVs does leave a noticeable imprint on
the optical spectrum as well, particularly for systems viewed at high
inclination. The strongest optical wind-formed recombinations lines are \ha\ and
\heiiopt\, although the outflow also produces emission in the Ly-$\alpha$
and \heiiuv\ UV lines. However, all of the optical lines produced by
this default wind model are double-peaked, and the effect of the wind
on the Balmer jump is modest. We carry out a limited exploration of
the disk wind parameter space, focusing on those parameters that affect the
density near the base of the wind. This shows that higher-density
outflows can produce both single-peaked emission lines and a
recombination continuum sufficient to fill in the Balmer
jump, although such models also tend to predict stronger-than-observed 
collisionally excited UV emission lines. Our main conclusion is that
disk wind emission is likely to have a significant impact on the
optical spectra of high-state CVs. 
{\bf JM: This abstract is a few lines longer than I would like- 
I think something punchier is required.}
\end{abstract}

@@

Much less is known about the effect of these outflows on the optical
spectra of high-state CVs. These spectra are typically characterized
by H and He emissions superposed on a blue continuum. In many
cases, and particularly in the SW~Sex subclass of NLs
\citep{HSK86,DR95}, these lines are single-peaked. This is contrary to
theoretical expectations for lines formed in accretion disks, which
are predicted to be double-peaked \citep{smak1981, hornemarsh1986}. 
{\em Low-state} CVs (dwarf novae in quiescence) can, in fact,
exhibit such double-peaked lines \citep{marshhorne1990}. 
{\bf JM: I've included a reference to a paper on IP Peg here
which shows the transition from double peaked lines in 
quiescence and a contrast to outburst spectra. I'm not sure
of general trends in DNe. A lot seem to have weak emission cores
embedded in absorption.}

@@

\subsection{Ionization and Excitation: ``Simple Atoms''}
\label{simpleatoms}

@@

\begin{figure}
\centering
\includegraphics[width=0.5\textwidth]{figures/fig_caseb_tardis.eps}
\caption{
{\sl Top Panel:} ``Case B'' Balmer decrements computed 
with \textsc{Python} compared to analytic calculations
by Seaton (1959). Both calculations are calculated at $T_e=10,000$K.
(see Osterbrock 1989 for a discussion of this commonly used approximation).
{\sl Bottom Panel:}  a comparison of Helium I level populations (the most complex ion we currently 
treat as a macro-atom) between \py~and \tar~models. 
The calculation is conducted with physical parameters of $n_e=5.96\times10^4$~cm$^{-3}$,
$T_e=30,600$K, $T_R=43,482$K and $W=9.65\times10^{-5}$. 
Considering the two codes use different atomic data and \textsc{Tardis,} unlike \textsc{Python,} currently has a complete treatment of collisions between 
radiatively forbidden transitions, the factor of 
$<2$ agreement is encouraging. 
}
\label{tests}
\end{figure}

\nocite{osterbrock}
\nocite{seaton1959}

@@

\subsection{Wind Geometry and Kinematics}
\label{kinematics}

@@

The density at position $(r,z)$ in the wind, $\rho(r,z)$, is
calculated from the mass continuity equation, yielding
\begin{equation}
\rho(r,z) = \frac{r}{r_0} \frac{dr}{dr_0} \frac{\phi(r_0)}{v_z(r,z)}.
\label{density}
\end{equation}
Here, $\phi(r_O)$ is the mass-loss rate per unit area at $r_0$, 
and $v_z$ is the vertical velocity component. We take $\phi(r_0)$ to be
uniform across the wind-launching area of the disk and normalize it by
matching its integral to the user-specified total mass-loss rate,
$\dot{M}_{wind}$. 
{\bf CK: I'm still not sure I understand the formula
  SV93 use to define $\phi$, and hence that this description is
  actually strictly correct.... Not sure what to do about that
  though...}

\subsection{Sources and Sinks of Radiation}
\label{radsources}

@@

It is possible to include radiation from a boundary layer (BL) between
the disk and the WD. In \py, the BL is described as
a blackbody with a user-specified effective temperature and
luminosity. In the models presented here, we have followed LK02 in setting
the BL luminosity to zero. However, we have confirmed that the addition of a
BL with $L_{BL} = 0.5 L_{acc}$ and temperatures in the range $80{\rm
kK} \leq T_{BL} \leq 200{\rm kK}$ would not change any of our main
conclusions. 
{\bf CK: Is this true? Do we want to say anything else, e.g. that we'll
talk about this in a future paper? JM: This is true. Anything up to 200kK
still has optical lines. The He II behaviour is not monotnic. CIV drops
gradually and gets a little overionized- but this is a separate topic really.
The issue of clumping and BL and their affect on ionization could be a 
future paper.}

@@

\begin{figure}
\centering
\includegraphics[width=0.45\textwidth]{figures/sed_figure.eps}
\caption{
The spectral energy distribution (SED) of the 
White Dwarf (dashed line) and blackbody accretion
disk used in the ionization calculation (black solid line).
The positions of the Hydrogen and Helium ionization edges 
are marked with vertical lines.
{\bf JM: do we want to show the stellar atmosphere spectrum here?
we know it looks a bit crap around the He II edge, so perhaps not...
 }
 }
\label{sed}
\end{figure}

@@

\begin{table}
\centering
\begin{tabular}{p{3cm}p{4cm}}
Benchmark Model (Optimized for the UV Band) \\
\hline Free Parameters 	&	 Value \\ 
\hline \hline 
$M_{WD}$ 	 &	 $0.8 M_{\odot}$ \\ 
$R_{WD}$ 	 &	 $7\times10^{8}$cm\\ 
$T_{WD}$ 	 &	 $40,000$K\\
$M_{2}$ 	 &	 $0.4 M_{\odot}$ \\ 
$\dot{M}_{acc}$ 	 &	 $10^{-8}~M_{\odot}yr^{-1}$\\ 
$\dot{M}_{wind}$  &	$10^{-9}~M_{\odot}yr^{-1}$å\\ 
$r_{min}$ 	&	 $4 R_{WD}$\\ 
$r_{max}$ 	&	 $12 R_{WD}$ \\ 
$\theta_{min}$ 	&	 $20.0^{\circ}$ \\ 
$\theta_{max}$ 	&	 $65.0^{\circ}$ \\ 
$\gamma$ 	&	 $1$ \\ 
$v_{\infty}$ 	&	 $3v_{esc}$ \\ 
%%$R_v$ 	        &	 $7\times10^{10}$cm \\ 
$R_v$ 	        &	 $100 R_{WD}$ \\ 
$\alpha$ 	&	 $1.5$ \\
\end{tabular}
\centering
\caption{
Parameters used for the geometry and kinematics of the benchmark CV model.
{\bf JM: this needs some stuff adding, as is mentioned in certain places}}
\label{wind_param}
\end{table}

\section{A Benchmark Disk Wind Model}
\label{modela}

@@

\subsection{Physical Structure and Ionization State}
\label{modela_ionization}

Figure~\ref{wind} shows the physical and ionization structure 
of the benchmark disk wind model. There is an obvious drop-off in density
and temperature with distance away from the disk, so any line
formation process that scales as $n_e^2$ - i.e. recombination and
collisionally excited emission - should be expected to operate
primarily in the dense base of the outflow. Moreover, a comparison of
the rotational and poloidal velocity fields shows that rotation
dominates in the near-disk regime, while outflow dominates further out
in the wind.

The ionization equation used in the ``simple atom'' approach used in
LK02 (see section~\ref{simpleatoms}) should be a reasonable approximation to
the photoionization equilibrium in the benchmark wind model. Even
though the macro-atom treatment of H and He does affect the 
computation of the overall ionization equilibrium, we would expect the
resulting ionization state of the wind to be quite similar to that
found in LK02. The bottom panels in figure~\ref{wind} confirm that this
is the case. In particular, Helium is fully ionized
throughout most of the outflow, except for a small region near the
base of the wind, which is shielded from the photons produced by the
hot inner disk. In line with the results shown in LK02, we also find
that C\textsc{iv} is the dominant Carbon ion throughout the wind,
resulting in a substantial absorbing column across a large range of
velocities. As we shall see, this produces the broad, deep and
blue-shifted C\textsc{iv}~$\lambda 1550{\rm \AA}$ absorption line that
is usually the most prominent wind-formed feature in the UV spectra of
low-inclination nova-like CVs.

\subsection{Synthetic Spectra}
\label{modela_spectrum}

We begin by verifying that the benchmark model still produces UV
spectra that resemble those observed in CVs. We do expect this to be
the case, since the ionization state of the wind has not changed
significantly from that computed in LK02 (see section~\ref{modela_ionization}). 
The left column of panels in figure~\ref{spec} shows that this expectation
is met: all of the strong metal resonance
lines - notably N\textsc{v}~$\lambda 1240{\rm \AA}$,
Si\textsc{iv}~$\lambda 1400{\rm \AA}$ and C\textsc{iv}~$\lambda
1550{\rm \AA}$ - are present and exhibit clear P-Cygni profiles
at intermediate inclinations. In addition, however, we now also find
that the wind produces significant Lyman-$\alpha$ and
He~\textsc{ii}~$\lambda1640{\rm \AA}$ emission lines.

@@

Figure~\ref{spec} shows the corresponding optical spectra produced for
the benchmark model, and these do exhibit some emission lines
associated with H and He. Since the emissivity of these recombination 
features scales as $n_e^2$, they are formed almost entirely in the 
dense base of the wind, just above the accretion disk. Here, the
velocity field of the wind is still dominated by rotation, rather than
outflow, which accounts for the double-peaked shape of the lines. In
principle, lines formed in this region can still be single peaked,
since the existence of a poloidal velocity {\em gradient} changes the
local escape probabilities (MC96). However, as
demonstrated explicitly in section~\ref{modelb}, the line opacity in our
benchmark model is not high enough for this radiative transfer effect
to dominate the line shapes.

@@

\section{The Dependence of Wind Features on Outflow Parameters}
\label{modelb}

The results of section~\ref{modela} show that wind models that reproduce
the UV spectral features of CVs are likely to leave a clear imprint on
the optical spectrum. It is therefore important to understand the
sensitivity of optical (and UV) wind-formed features to the basic
properties of the outflow. In particular, there are two key questions
we would like to address: (i) under what conditions can disk winds
produce {\em single-peaked} Balmer emission lines (MC96); (ii) can the wind-formed recombination continuum completely
fill in the disk's Balmer absorption edge for reasonable outflow
parameters?

@@

The simplest way to simultaneously affect the density in the wind as
well as the velocity gradients is by modifying the poloidal velocity
law. We have therefore conducted a exploratory sensitivity study in
which we focus on just two kinematic variables (section~\ref{kinematics}):

\begin{itemize}
 	\item the acceleration length, $R_v$, which controls the
        distance over which the wind accelerates to $\frac{1}{2}v_{\infty}$;
 	\item the acceleration exponent, $\alpha$, which controls the rate 
 	at which the poloidal velocity changes near $R_v$.
\end{itemize} 
The general behaviour we might expect is that outflows with denser
regions near the wind base -- i.e. winds with larger $R_{v}$ and/or
larger $\alpha$ -- will produce stronger optical emission signatures. 
However, this behaviour may be moderated by the effect of the increasing
optical depth through this region.

\subsection{Optical Wind Signatures}
\label{modelb_opt}

Figure~\ref{halpha} shows how the H$\alpha$ and 
He~{\sc ii}~$\lambda 
4686$ lines change with the kinematics of the wind for an inclination
of $80^\circ$. The width and strength of He~{\sc ii} clearly depend on
both $R_v$ and $\alpha$, but the overall line profile shape is always
double-peaked. This contrasts with the behaviour of H$\alpha$, which
becomes single-peaked in our densest models. The transition from
double-peaked to single-peaked occurs when the line becomes optically
thick. This reproduces the behaviour predicted by \cite{MC96}, who
showed that the poloidal velocity gradient in the wind can modify the
shape of optically thick emission lines even in the near-disk regime,
where the poloidal velocity itself is much smaller than the rotational
velocity. This has the effect of suppressing the line wings and
can therefore lead to the production of a single-peaked line profile
shape.

@@

Two other features are worth noting in this part of the
spectrum. First, the collisionally excited Ca~{\sc ii} H\&K 
emission lines at 3969\AA 
and (especially) 3934\AA become quite prominent in our densest
models. Second, {\em all} of our models predict a detectable
He~\textsc{ii} recombination line at 3202\AA. This is the Helium
equivalent of Paschen~$\beta$ and should be expected in all systems that
feature a strong He~\textsc{ii}~4686\AA\ line (the Helium
equivalent of Paschen~$\alpha$). This line is unfamiliar observationally,
but only because it lies bluewards of the atmospheric cut-off, but
also redwards of most ultraviolet spectra. 
{\bf JM: we should we reference something here for the 3202 line. 
I blieve CK found a paper which shows it in an IP? I'll try and find it.}

\begin{figure} %fullpage
\mbox{
\subfigure{\includegraphics[width=0.5\textwidth]{figures/3by3_grid_alpha.eps}}
\quad
\subfigure{\includegraphics[width=0.5\textwidth]{figures/3by3_grid_4686.eps}}   
}
\caption{
Line profiles plotted in veliocity space 
for \ha\ (left) and \heiiopt\ (right) with varying kinematic 
properties, computed for an inclination of $80^\circ$.
The benchmark model and the improved optical
model described in section 6 are labeled as A and B respectively. 
The $x$-axis limits correspond to the Keplerian velocity at 
$4R_{WD}$, the inner edge of the wind.
The parameter that has been altered from the benchmark model 
is marked in each column and row. Models towards the bottom right
are have more slowly accelerating winds and denser wind bases. 
{\bf JM: Need to discuss the `single-peaked-ness' of line profiles
in our telecon.}
}
\label{halpha}
\end{figure} %fullpage

\begin{figure} %fullpage
\centering
\includegraphics[width=1.0\textwidth]{figures/3by3_grid_balmer.eps}  
%%\includegraphics[width=0.5\textwidth]{figures/test.eps}
\caption{
The evolution of the Balmer jump with varying kinematic 
properties, computed for an inclination of $80^\circ$. 
The panels are organised in the same way as figure~\ref{halpha}.
% The panels are ordered in the same way as figure~\ref{halpha}. In general, increasing densities
% lead to a Balmer jump in emission, although as in the case of H$\alpha$ there is a transition
% when the wind becomes a neutral absorbing column.
}
\label{jump}
\end{figure} %fullpage

\subsection{Ultraviolet Wind Signatures}
\label{modelb_uv}

The dependence of the ultraviolet C~{\sc iv} and He~{\sc ii} lines on
the same kinematic wind parameters is illustrated in figures~\ref{uvlines}
and~\ref{uvlines}. As before, we focus on the predicted behaviour at $i =
80^\circ$. In our benchmark model, the C~{\sc iv} emission
line seen at high inclinations is almost entirely produced via
resonance scattering. The angle-averaged flux integrated over the
entire line profile - accounting correctly for the blue-shifted
absorption component seen at low inclinations - would be
approximately zero. However, as we increase the density and extent of
the slow-moving part of the outflow near the disk plane, collisionally
excited emission starts to contribute significantly to the C~{\sc iv}
line profiles. This explains the much stronger line-to-continuum
contrast seen in the $\alpha = 1.5$, $R_v = 2\times 10^{11}$~cm model,
for example. For any given inclination, the trend is not necessarily
monotonic, however, because the escape probabilities from the
line-forming region can be highly anisotropic.  
{\bf CK: I *think* this is the basic reason, but (a) can you check
  this, James; (b) is it enough to just point this out?; (c) do we
  understand why CIV is single-peaked but HeII, for example, is double-peaked?
  JM: There are still a few tests I want to do here- we can discuss in the telecon.
  I believe He II v CIV shape is just due to CIV being produced in an extended region,
  and generally scattered, whereas He II is produced close to the disk, and there's probably
  also a line opacity effect.
  I can check this.}

@@

\begin{figure} %fullpage
\mbox{
\subfigure{\includegraphics[width=0.5\textwidth]{figures/3by3_grid_uv.eps}}
\quad
\subfigure{\includegraphics[width=0.5\textwidth]{figures/3by3_grid_1640.eps}}   
}
%%\includegraphics[width=0.5\textwidth]{figures/test.eps}
\caption{
Line profiles for \civ\ (left) and \heiiuv\  (right) with varying kinematic 
properties, computed for an inclination of $80^\circ$. 
The panels are organised in the same way as figure~\ref{halpha}.
{\bf JM: I need to adjust the smoothing here so for all plots I smooth
in terms of Wavelength rather than bins. Linked to single-peaked discussion. 
Note also CK may want a Lyman alpha plot, which I might put here}
% The panels are ordered in the same way as figure~\ref{halpha}. In general, increasing densities
% lead to a Balmer jump in emission, although as in the case of H$\alpha$ there is a transition
% when the wind becomes a neutral absorbing column.
}
\label{uvlines}
\end{figure}

@@

However, perhaps the most serious problem for this optically optimized
model is associated with its predicted UV spectra. In order to
generate strong optical wind signatures, we have adopted wind
parameters that lead to very high densities at the base of the wind
($n_e\sim10^{13}-10^{14}$ in the main emission region). This produces
the desired optical recombination emission, but also increases the
role of collisional excitation in the formation of the UV resonance
lines. This explains the pronounced increase in the emission component 
of the C\textsc{iv} 1550\AA resonance line, for example, relative to
what was seen in the benchmark model (compare figures~\ref{spec} and
\ref{uvoptb}). The strength of this component in the revised model --
especially at low and intermediate inclinations -- is probably
somewhat too high to be consistent with UV observations of high-state
CVs (REF).
